The State of Urbit:\\Eight Years After the Whitepaper

### 2024-04-24 mss.tex
Revise for final version.

While the whitepaper by its own admission presented itself as a ``work-in-progress presentation'' of a ``semi-closed alpha test'', the basic claims of the Urbit whitepaper are still true today.  Most of the intervening work over the past eight years has gone into improving robustness and better support for applications.  A capsule history of ``late early'' Urbit includes the following milestones:
\begin{itemize}
  \item  2015–2017:  Early vane explosion
  \item  2017–2019:  Hoon, Ford, Eyre, and Vere rewrites refine the system
  \item  2018–2019:  Azimuth and Ames rewrite refine the network
  \item  2018–2019:  Nock bytecode interpreter
  \item  2020:  Ford Fusion enables network continuity
  \item  2020:  Security audit on Ames
  \item  2021:  Azimuth Layer2
  \item  2021:  Software distribution introduced
  \item  2022–2024:  Software distribution maturation
  \item  2023–2024:  Network scaling
\end{itemize}

@@

Since 2016, Nock underwent a revision from Kelvin version 5K down to 4K.  This was a refinement rather than a deviation, mainly adding an opcode to manage tree mutation.  Hoon has undergone a number of changes but is certainly recognizable as the same language.

@@

\subsection{Clay ``Theory of a Desk''}

@@

The ``theory of a desk'' paradigm uses only the current snapshot of the desk to build any mark files needed to validate the files in the desk.  This was not technically impossible while Ford was a separate vane, but the asynchronous communication back and forth made it exceedingly difficult.  The new Ford shrank from 6,000 lines of code down to about 500, and it was much easier to reason about and debug.

@@

A major part of making the namespace more usable was to expose namespace queries over the network.  Designed in 2021 and deployed in 2023, the ``Fine''\footnote{Pronounced \emph{fee-NAY}.} remote scry network protocol was added to Urbit.  The first use cases were Clay, which uses it for downloading desks, and the Gall \lstinline[style=inlinecode]{%grow} namespace, described in more detail in a later section.

@@

Arvo has traditionally used ``depth-first'' ordering when processing ``moves'', i.e. messages passed between parts of the system.  Since each module in Arvo emits a list of moves whenever it runs, and then the moves are delivered one at a time to their destination modules for processing, at any given time there is a tree of unprocessed moves that have been emitted but not delivered.

@@

This means \lstinline[style=inlinecode]{%foo} can rely on \lstinline[style=inlinecode]{%bar} completely handling the first poke before \lstinline[style=inlinecode]{%baz} receives the second poke, even if \lstinline[style=inlinecode]{%bar} has to issue further commands in order to do so – at least as long as all the pokes are, one, domestic (in the same ship) and two, processable synchronously, i.e. no operations need to be spread across multiple Arvo events, such as waiting for a timer to elapse or waiting for a user to type input on the command line.  If both these conditions are met, then the poke to \lstinline[style=inlinecode]{%bar} has one view of the system's state, and the poke to \lstinline[style=inlinecode]{%baz} has a newer, updated view, but both views are consistent in the sense that for either one of them, \lstinline[style=inlinecode]{%bar}'s move processing has either not started at all or completely finished.

@@

The latest ``shrubbery'' (q.v.) experiment uses a hybrid move ordering: depth-first for local pokes within the same application, and breadth-first otherwise.

@@

The latest ``shrubbery'' experiment attempts a hybrid solution to get the best of both modalities.  A typical shrubbery ``app'' will be broken into a hierarchy of event handlers at various subpaths.  When sending a move ``downward'' into a subpath, the system will handle that move depth-first, since an app is presumed to know the characteristics of its sub-apps.  These moves will not need to be acknowledged, since they will be truly depth-first.  When passing a move ``upward'' or ``sideways'', i.e. outside of the sub-tree, that move will be appended to one global queue, to be handled breadth-first.

@@

\subsection{Azimuth: On-Chain PKI}

One major event that improved the level of continuity was moving Urbit's public key infrastructure ({\sc pki}) from an informal spreadsheet maintained by the Tlon Corporation to a set of smart contracts on the Ethereum blockchain.

@@

\sloppy
This allows a user to continue to use the address they purchased even if their Urbit loses data.  The newly written \lstinline[style=inlinecode]{%phoenix} app implements a system of encrypted backups, so that after a breach, a user can recover application data from the latest backup.

@@

The next major frontier in continuity will be ``Kelvin shims'', i.e. a kernel that can keep stale apps designed to work with previous kernels running (through a ``shim'' for a previous Kelvin version), not just apps designed to run at exactly the current version.  This is a thorny problem that the core development team experimented with in early 2023 but abandoned.

The hard part of Kelvin shimming turned out to be that the current version of the userspace/kernelspace {\sc api} is complex.  Taming that interface and localizing application-related state to make it easier to migrate will be critical for enabling Kelvin shims.  The shrubbery experiment is a step in this direction.

@@

In early 2023, the responsibility for core development transferred from the Tlon Corporation, which had managed it since 2013, to the newer Urbit Foundation, a nonprofit.  The Urbit Foundation has become the custodian of the Urbit kernel and ecosystem.  UF adopted and continued to modify Tlon's release workflow, including CI/CD (continuous integration/continuous deployment), multi-stage canary deployments, and repository management practices.  This process has more recently been extended to include a liaison period for each Kelvin update where organizations that use Urbit can use the prerelease software on their own infrastructure first and report any bugs to UF.

@@

In 2016, the Urbit network could barely handle a few hundred online ships.  In 2024, at 411 Kelvin, the network can easily handle many tens of thousands of online ships, if not more.  Packet forwarding is now stateless and does not run Nock.  Sponsor pinging now uses the {\sc stun} protocol, incurring no disk writes on sponsoring galaxies.

The Directed Messaging project will increase network throughput by roughly a factor of a thousand, fast enough to saturate a gigabit ethernet link.  The Ares project will increase the maximum persistent data storage by a factor of several thousand, up to 64 Terabytes.  Ares should also improve Nock execution speed dramatically, paving the way toward Nock being able to be compiled to machine code comparable to that emitted by a C compiler.

@@

Kernel and app upgrades are much more reliable.  The system runs out of memory much less often.  The network maintains connectivity much better.  Commonly used applications have fewer bugs.  Browser page loads are much faster and more reliable.  Individual continuity breaches cause fewer downstream bugs in the kernel and applications.  Arvo error handling has fewer known flaws.  Gall subscriptions underwent a major rearchitecture in 2022 to fix a queue desynchronization bug.  The Behn timer module got rewritten multiple times to shake out several different bugs.

@@

The Ares project, now close to initial deployment, includes a clean-slate rewrite of the Nock interpreter, using an all-new allocator, a novel static analysis technique called Subject Knowledge Analysis to achieve substantial Nock execution speed improvements, and a new persistence module called the Persistent Memory Arena ({\sc pma}), using a copy-on-write B+ tree to keep memory pages synchronized between disk and {\sc ram}.

@@

Fortunately, because Urbit supports structural sharing and Ford is a purely functional build system, a referentially transparent cache was added to Ford.  This cache does not know which desk a build is being run in, so it can cache and deduplicate builds across desks seamlessly without losing the logical static linking.

This means Urbit gets the best of both worlds: the sanity of static linking, and the memory deduplication of dynamic linking.

@@

  This had a few problems.  It allocated more Nock cells at runtime than needed.  It prevented the interpreter from optimizing in-place mutation of a noun with a reference count of one.  If no one else is looking, creating a mutant copy can be done by overwriting the noun in-place and nobody can tell on you.  APL interpreters, which have a similar memory model to Nock nouns, usually include this optimization, and Urbit's current runtime does as well.

@@

Hoon's recent changes have reduced the prevalence of the ``{\sc tmi} problem''.  Since Hoon's type system supports subtyping, it is common for code to fail to compile due to having ``too much information'' ({\sc tmi}) – sometimes a more specific type cannot be used in place of a more general one, namely when that type is in ``contravariant position''.  For example, if the type of some field in the Hoon subject was a list but got specialized in a conditional branch to be a non-null list, attempting to ``overwrite'' (really, create a mutant copy of) that field in the subject with a normal list, the compiler will throw a type error because it cannot guarantee that the normal list is not null.  If the compiler did not know the subject's list field was non-null, then overwriting it with some other list would just.work.  Almost always, the eventual result of the expression admits a null list, so the type error is an annoyance rather than an assurance.

Hoon type system changes have also included multiple attempts at resolving another difficulty with subtyping: ``wetness''.  A ``wet core'' is a Hoon core (pair of code and data, where the code is run using the whole core as the subject) with a ``sample'' (argument slot – the head of the ``data'' tail by convention in the type system) whose type can vary.  This is very closely related to the mathematical idea of parametric polymorphism.  Like in many languages (including Haskell), Hoon does not strictly guarantee ``parametricity'', but wetness is used to implement parametrically polymorphic functions and even type constructors (called ``mold builders'' in Hoon).

Wetness has an issue where the ``faces'' (names for particular slots within the tree) in the sample type in the definition of the wet core can differ from the faces in the sample type at any given call site.  Deciding which of those faces to use in the body of the code is a subtle, thorny problem.  The Hoon at the time of the whitepaper had one solution; another was written afterward, around 2017; a third was deployed in 2019.  This last one, called ``repainting'', is generally thought to be the right strategy, but its implementation remains unfinished in 2024 – it has an issue where the precedence of call-site vs. definition-site face names is not preserved under recursion.  This issue will need to be fixed before we can say the right answer to this issue has been achieved.

@@

Meanwhile, in userspace, the \lstinline[style=inlinecode]{%spider} agent was added, to run a new kind of userspace program: a \lstinline[style=inlinecode]{%thread}.  This is not a separate processor thread of execution, but rather a system of monadic IO based heavily on Haskell's IO monad.  Unlike agents, which are designed to allow a programmer to implement multiple long-lived asynchronous operations concurrently using non-blocking IO, a thread is designed for building a linear sequence of asynchronous operations, more akin to a coroutine.

@@

The \lstinline[style=inlinecode]{.^} dotket rune no longer blocks until complete.  Because the agent model expects non-blocking I/O, a single dotket could make an entire agent stuck if it can't be completed.  While a lot of work has gone into fleshing out the scry namespace, core developers are now entertaining the notion that the expectation of synchronous access to the namespace might not be worth it for agents.  An Urbit Improvement Proposal has been submitted that would remove the \lstinline[style=inlinecode]{.^} dotket rune in favor of a move-based interface to the namespace (see pp. TODO in this issue).

@@

Significant work has gone into improving the runtime's implementation of uniform persistence:
\begin{itemize}
  \item  demand paging to allow storage of more data on disk than fits in {\sc ram}
  \item  a guard page to efficiently prevent stack smashing
  \item  a snapshot migration framework to facilitate upgrades
  \item  separating ephemeral and persistent state to avoid extraneous disk writes
  \item  using the {\sc lmdb} database for the event log, instead of a custom binary file
  \item  the Ares project's Persistent Memory Arena ({\sc pma}), a general-purpose single-level store that can efficiently manage the persistence and paging for 64 terabytes of data, using a copy-on-write B+ tree
  \item  the ``epoch system'', which broke the event log into epochs, allowing event log rollover and truncation
  \item  persistent Nock caches, allowing arbitrary computations to be memoized in the runtime in a way that persists across Arvo events and Unix process restarts
\end{itemize}

@@

The canonical pronunciations for syllables was changed in 2019, then reverted due to effectively unanimous demand.  The one modification that persisted was the change of \lstinline[style=inlinecode]{;} from `sem' to `mic', since `sem' sounded too close to 'cen' (\lstinline[style=inlinecode]{%}).

@@

Instead, the encrypted remote system has its own separate concept of a ``security domain'' as a path prefix, and the kernel performs a dynamic query when a peer asks for access to private data with that prefix.  While the performance is not a direct comparison with the security mask solution, the dynamic query can typically be implemented in logarithmic time by the application, e.g. a binary search through a Hoon \lstinline[style=inlinecode]{$map} data structure (a treap).  For more complex permissions models, this approach should scale much better.

@@

In practice, the Ames protocol has changed a few times since the whitepaper, but the cryptosuite has not.  Allowing the cryptosuite to continue to upgrade after Kelvin zero will probably still need to be implemented, but we have not yet done so.

@@

\sloppy
This requires another vaporware warning, but one could imagine a userspace application sending three poke commands to `~zod` by binding the commands to three paths like \lstinline[style=inlinecode]{/poke/~zod/foo/1}, \lstinline[style=inlinecode]{/poke/~zod/foo/2}, and \lstinline[style=inlinecode]{/poke/~zod/foo/3}, into a subtree of its own scry namespace.  The kernel would ensure \patp{zod} is notified of those bindings.  \patp{zod} would download the values at those paths and notify its \lstinline[style=inlinecode]{%foo} application.  The app would interpret the values as a sequence of poke commands, to be applied once and in order.  Once it processes each poke, it would bind an acknowledgment value to a predetermined path in its own namespace, such as \lstinline[style=inlinecode]{/ack/~nec/1}.

\paragraph{9.7 Permanent Applications}  Applications are generally no longer considered permanent.  The current Gall model lets a user suspend an application and revive it later.  Unfortunately the current model is known to be unsound: ``signs'', i.e. responses to outgoing commands or queries, to a Gall agent are dropped while the agent is suspended, leading to an inconsistent state.

@@

Something namespace-oriented, like the example with poking \patp{zod} by binding paths in our local namespace, could be more amenable to maintaining exactly-once delivery across suspension and revival without the kernel needing to maintain an unbounded queue of acks to deliver to the agent.  It could also help with recoverability after a breach, since storing userspace data with absolute references (i.e. scry paths) allows for more seamless social recovery – if any peer still has the scry binding, the breached ship could download and reinstate it.  The importance of immutable absolute references for data is one of the key insights leading to the ``shrubbery'' noun-maximalist way of thinking.

@@

App publishers can push down source code to their users for the next version of their app, intended to be built and run using the next Kelvin version of the kernel.  When the user's ship receives the kernel's source code for the new Kelvin, it upgrades the apps and the kernel atomically with each other, ensuring all apps (soon: only apps marked as ``essential'') update succesfully – if an app fails to upgrade to the new Kelvin version, the Arvo event is aborted, rolling back the kernel upgrade and any other app upgrades that had happened during the event.

Eventually, the kernel should be able to keep agents designed for older kernels running.  We call this idea a ``Kelvin shim''.  A project along these lines was attempted at the start of 2023, but the current userspace model turns out to be to be too tightly coupled to the kernel for shimming to be feasible.  This experience has informed our next generation of ideas for the kernelspace/userspace interface.

@@

There are more use cases for jammed nouns now.  Portable Arvo snapshots are jammed nouns.  Interprocess communication between the two Vere processes uses jammed nouns for its messages.  The Khan and Lick vanes also use jammed nouns for IPC, and Eyre, the {\sc http} vane, now supports {\sc http} subscriptions using jammed nouns in addition to {\sc json}.

@@

\printbibliography
\end{document}

>>>>

While the whitepaper by its own admission presented itself as a ``work-in-progress presentation'' of a ``semi-closed alpha test'', the basic claims of the Urbit whitepaper are still true today.  Most of the intervening work over the past eight years has gone into improving robustness and better support for applications.  A capsule history of ``late early'' Urbit includes the following milestones:
\begin{itemize}
  \item  2015–2017:  Early vane explosion
  \item  2017–2019:  Hoon, Ford, Eyre, and Vere rewrites refine the system
  \item  2018–2019:  Azimuth and Ames rewrite refine the network stack
  \item  2018–2019:  Nock bytecode interpreter
  \item  2020:  Ford Fusion enables network continuity
  \item  2020:  Security audit on Ames
  \item  2021:  Azimuth Layer2
  \item  2021:  Software distribution introduced
  \item  2022–2024:  Software distribution maturation
  \item  2023–2024:  Network scaling
\end{itemize}

@@

Since 2016, Nock has undergone a revision from kelvin version 5K down to 4K.  This was a refinement rather than a deviation, mainly adding an opcode to manage tree mutation (see appendix for details).  Hoon has undergone a number of changes but is certainly recognizable as the same language.

@@

\subsection{Clay “Theory of a Desk”}

@@

The ``theory of a desk'' paradigm uses only the current snapshot of the desk to build any mark files needed to validate the files in the desk.  This was not technically impossible while Ford was a separate vane, but the asynchronous communication back and forth made it exceedingly difficult.  The new Ford shrank from 6,000 lines of code down to about 500, and became much easier to reason about and debug.

@@

A major part of making the namespace more usablae was to expose namespace queries over the network.  Designed in 2021 and deployed in 2023, the ``Fine''\footnote{Pronounced \emph{fee-NAY}, after French cartographer and mathematician Oronce Fine.} remote scry network protocol was added to Urbit.  The first use cases were Clay, which uses it for downloading desks, and the Gall \lstinline[style=inlinecode]{%grow} namespace, described in more detail in a later section.

@@

Arvo has traditionally used ``depth-first'' ordering when processing ``moves'', i.\,e.\ messages passed between parts of the system.  Since each module in Arvo emits a list of moves whenever it runs, and then the moves are delivered one at a time to their destination modules for processing, at any given time there is a tree of unprocessed moves that have been emitted but not delivered.

@@

This means \lstinline[style=inlinecode]{%foo} can rely on \lstinline[style=inlinecode]{%bar} completely handling the first poke before \lstinline[style=inlinecode]{%baz} receives the second poke, even if \lstinline[style=inlinecode]{%bar} has to issue further commands in order to do so – at least as long as all the pokes are, one, domestic (in the same ship) and two, processable synchronously, i.\,e.\ no operations need to be spread across multiple Arvo events, such as waiting for a timer to elapse or waiting for a user to type input on the command line.  If both these conditions are met, then the poke to \lstinline[style=inlinecode]{%bar} has one view of the system's state, and the poke to \lstinline[style=inlinecode]{%baz} has a newer, updated view, but both views are consistent in the sense that for either one of them, \lstinline[style=inlinecode]{%bar}'s move processing has either not started at all or completely finished.

@@

The latest ``shrubbery'' (q.\,v.)\ experiment uses a hybrid move ordering: depth-first for local pokes within the same application, and breadth-first otherwise.

@@

The latest ``shrubbery'' experiment attempts a hybrid solution to get the best of both modalities.  A typical shrubbery ``app'' will be broken into a hierarchy of event handlers at various subpaths.  When sending a move ``downward'' into a subpath, the system will handle that move depth-first, since an app is presumed to know the characteristics of its sub-apps.  These moves will not need to be acknowledged, since they will be truly depth-first.  When passing a move ``upward'' or ``sideways'', i.\,e.\ outside of the sub-tree, that move will be appended to one global queue, to be handled breadth-first.

@@

\subsection[Azimuth:  On-Chain {\sc pki} (2019)]{Azimuth: On-Chain PKI (2019)}

One major event that improved the level of continuity was moving Urbit's public key infrastructure ({\sc pki}) from an informal spreadsheet maintained by the Tlon Corporation to a set of smart contracts on the Ethereum blockchain \citep{WolfePauly2019}.

@@

\sloppy
This allows a user to continue to use the address they purchased even if their Urbit loses data.  The newly written \lstinline[style=inlinecode]{%phoenix} app implements a system of encrypted backups, so that after a breach, a user can recover application data from the latest backup \citep{Wilson2024}.

@@

The next major frontier in continuity will be ``kelvin shims'', i.\,e.\ a kernel that can keep stale apps designed to work with previous kernels running (through a ``shim'' for a previous kelvin version), not just apps designed to run at exactly the current version.  This is a thorny problem that the core development team experimented with in early 2023 but abandoned.

The hard part of kelvin shimming turned out to be that the current version of the userspace/kernelspace {\sc api} is complex.  Taming that interface and localizing application-related state to make it easier to migrate will be critical for enabling kelvin shims.  The shrubbery experiment is a step in this direction.

@@

In early 2023, the responsibility for core development transferred from the Tlon Corporation, which had managed it since 2013, to the newer Urbit Foundation, a nonprofit.  The Urbit Foundation has become the custodian of the Urbit kernel and ecosystem.  UF adopted and continued to modify Tlon's release workflow, including CI/CD (continuous integration/continuous deployment), multi-stage canary deployments, and repository management practices.  This process has more recently been extended to include a liaison period for each kelvin update where organizations that use Urbit can use the prerelease software on their own infrastructure first and report any bugs to UF.

@@

In 2016, the Urbit network could barely handle a few hundred online ships.  In 2024, at 411 kelvin, the network can easily handle many tens of thousands of online ships, if not more.  Packet forwarding is now stateless and does not run Nock.  Sponsor pinging now uses the {\sc stun} protocol, incurring no disk writes on sponsoring galaxies.

The Directed Messaging project will increase network throughput by roughly a factor of a thousand, fast enough to saturate a gigabit ethernet link.  The Ares project will increase the maximum persistent data storage by a factor of several thousand, up to 64 terabytes.  Ares should also improve Nock execution speed dramatically, paving the way toward Nock being able to be compiled to machine code comparable to that emitted by a C compiler.

@@

Kernel and app upgrades are much more reliable.  The system runs out of memory much less often.  The network maintains connectivity much better.  Commonly used applications have fewer bugs.  Browser page loads are much faster and more reliable.  Individual continuity breaches cause fewer downstream bugs in the kernel and applications.  Arvo error handling has fewer known flaws.  Gall subscriptions underwent a major rearchitecture in 2022 to fix a queue desynchronization bug.  The Behn timer module was rewritten multiple times to shake out several different bugs.

@@

The Ares project, now close to initial deployment, includes a clean-slate rewrite of the Nock interpreter, using an all-new allocator, a novel static analysis technique called Subject Knowledge Analysis to achieve substantial Nock execution speed improvements, and a new persistence module called the Persistent Memory Arena, using a copy-on-write B+ tree to keep memory pages synchronized between disk and {\sc ram}.

@@

Fortunately, because Urbit supports structural sharing and Ford is a purely functional build system, a referentially transparent cache was added to Ford.  This cache does not know which desk a build is being run in, so it can cache and deduplicate builds across desks seamlessly without losing the logical static linking (see pp. 75–82 in this issue).

Urbit thus gets the best of both worlds: the sanity of static linking and the memory deduplication of dynamic linking.

@@

  This had a few problems.  It allocated more Nock cells at runtime than needed.  It prevented the interpreter from optimizing in-place mutation of a noun with a reference count of one.  If no one else is looking, creating a mutant copy can be done by overwriting the noun in-place and nobody can tell on you.  {\sc apl} interpreters, which have a similar memory model to Nock nouns, usually include this optimization, and Urbit's current runtime does as well.

@@

Hoon's recent changes have reduced the prevalence of the ``{\sc tmi} problem''.  Since Hoon's type system supports subtyping, it is common for code to fail to compile due to having ``too much information'' ({\sc tmi}) – sometimes a more specific type cannot be used in place of a more general one, namely when that type is in ``contravariant position''.  For example, if the type of some field in the Hoon subject was a list but was specialized in a conditional branch to be a non-null list, attempting to ``overwrite'' (really, create a mutant copy of) that field in the subject with a normal list, the compiler will throw a type error because it cannot guarantee that the normal list is not null.  If the compiler did not know the subject's list field was non-null, then overwriting it with some other list would ``just work''.  Almost always, the eventual result of the expression admits a null list, so the type error is an annoyance rather than an assurance.

Hoon type system changes have also included multiple attempts at resolving another difficulty with subtyping: ``wetness''.  A ``wet core'' is a Hoon core (pair of code and data, where the code is run using the whole core as the subject) with a ``sample'' (argument slot – the head of the ``data'' tail by convention in the type system) whose type can vary.  This is very closely related to the mathematical idea of parametric polymorphism.  Like some other languages (including Haskell), Hoon does not strictly guarantee ``parametricity'', but wetness is used to implement parametrically polymorphic functions and even type constructors (called ``mold builders'' in Hoon).

Wetness has an issue where the ``faces'' (names for particular slots within the tree) in the sample type in the definition of the wet core can differ from the faces in the sample type at any given call site.  Deciding which of those faces to use in the body of the code is a subtle, thorny problem.  The Hoon at the time of the whitepaper employed one solution; another was written afterward, around 2017; a third was deployed in 2019.  This last revision, called ``repainting'', is generally thought to be the right strategy, but its implementation remains unfinished in 2024 – it has an issue where the precedence of call-site vs. definition-site face names is not preserved under recursion.  This issue will need to be fixed before we can say the right answer to this issue has been achieved.

@@

Meanwhile, in userspace, the \lstinline[style=inlinecode]{%spider} agent was added, to run a new kind of userspace program: a \lstinline[style=inlinecode]{%thread}.  This is not a separate processor thread of execution, but rather a system of monadic I/O based heavily on Haskell's I/O monad.  Unlike agents, which are designed to allow a programmer to implement multiple long-lived asynchronous operations concurrently using non-blocking I/O, a thread is designed for building a linear sequence of asynchronous operations, more akin to a coroutine.

@@

The \lstinline[style=inlinecode]{.^} dotket rune no longer blocks until complete.  Because the agent model expects non-blocking I/O, a single dotket could make an entire agent stuck if it can't be completed.  While a lot of work has gone into fleshing out the scry namespace, core developers are now entertaining the notion that the expectation of synchronous access to the namespace might not be worth it for agents.  An Urbit Improvement Proposal has been submitted that would remove the \lstinline[style=inlinecode]{.^} dotket rune in favor of a move-based interface to the namespace (see pp. 47–58 in this issue).

@@

Significant work has gone into improving the runtime's implementation of uniform persistence:
\begin{itemize}
  \item  demand paging to allow storage of more data on disk than fits in {\sc ram};
  \item  a guard page to efficiently prevent stack smashing;
  \item  a snapshot migration framework to facilitate upgrades;
  \item  separating ephemeral and persistent state to avoid extraneous disk writes;
  \item  using the {\sc lmdb} database for the event log, instead of a custom binary file;
  \item  the Ares project's Persistent Memory Arena ({\sc pma}), a general-purpose single-level store that can efficiently manage the persistence and paging for 64 terabytes of data, using a copy-on-write B+ tree;
  \item  the ``epoch system'', which broke the event log into epochs, allowing event log rollover and truncation;
  \item  persistent Nock caches, allowing arbitrary computations to be memoized in the runtime in a way that persists across Arvo events and Unix process restarts.
\end{itemize}

@@

The canonical pronunciations for syllables was changed in 2019, then reverted due to effectively unanimous demand.  The one modification that persisted was the change of \lstinline[style=inlinecode]{;} from `sem' to `mic', since `sem' sounded too close to `cen' (\lstinline[style=inlinecode]{%}).

@@

Instead, the encrypted remote system has its own separate concept of a ``security domain'' as a path prefix, and the kernel performs a dynamic query when a peer asks for access to private data with that prefix.  While the performance is not a direct comparison with the security mask solution, the dynamic query can typically be implemented in logarithmic time by the application, e.\,g.\ a binary search through a Hoon \lstinline[style=inlinecode]{$map} data structure (a treap).  For more complex permissions models, this approach should scale much better.

@@

In practice, the Ames protocol has changed a few times since the whitepaper, but the cryptosuite has not.  Allowing the cryptosuite to continue to upgrade after kelvin zero will probably still need to be implemented, but we have not yet done so.

@@

\sloppy
This requires another vaporware warning, but one could imagine a userspace application sending three poke commands to \patp{zod} by binding the commands to three paths like \lstinline[style=inlinecode]{/poke/~zod/foo/1}, \lstinline[style=inlinecode]{/poke/~zod/foo/2}, and \lstinline[style=inlinecode]{/poke/~zod/foo/3}, into a subtree of its own scry namespace.  The kernel would ensure \patp{zod} is notified of those bindings.  \patp{zod} would download the values at those paths and notify its \lstinline[style=inlinecode]{%foo} application.  The app would interpret the values as a sequence of poke commands, to be applied once and in order.  Once it processes each poke, it would bind an acknowledgment value to a predetermined path in its own namespace, such as \lstinline[style=inlinecode]{/ack/~nec/1}.

\paragraph{9.7 Permanent Applications}  Applications are generally no longer considered permanent.  The current Gall model lets a user suspend an application and revive it later.  Unfortunately the current model is known to be unsound: ``signs'', i.\,e.\ responses to outgoing commands or queries, to a Gall agent are dropped while the agent is suspended, leading to an inconsistent state.

@@

Something namespace-oriented, like the example with poking \patp{zod} by binding paths in our local namespace, could be more amenable to maintaining exactly-once delivery across suspension and revival without the kernel needing to maintain an unbounded queue of acks to deliver to the agent.  It could also help with recoverability after a breach, since storing userspace data with absolute references (i.\,e.\ scry paths) allows for more seamless social recovery – if any peer still has the scry binding, the breached ship could download and reinstate it.  The importance of immutable absolute references for data is one of the key insights leading to the ``shrubbery'' noun-maximalist way of thinking.

@@

App publishers can push down source code to their users for the next version of their app, intended to be built and run using the next kelvin version of the kernel.  When the user's ship receives the kernel's source code for the new kelvin, it upgrades the apps and the kernel atomically with each other, ensuring all apps (soon: only apps marked as ``essential'') update succesfully – if an app fails to upgrade to the new kelvin version, the Arvo event is aborted, rolling back the kernel upgrade and any other app upgrades that had happened during the event.

Eventually, the kernel should be able to keep agents designed for older kernels running.  We call this idea a ``kelvin shim''.  A project along these lines was attempted at the start of 2023, but the current userspace model turns out to be to be too tightly coupled to the kernel for shimming to be feasible.  This experience has informed our next generation of ideas for the kernelspace/userspace interface.

@@

There are more use cases for jammed nouns now.  Portable Arvo snapshots are jammed nouns.  Interprocess communication between the two Vere processes uses jammed nouns for its messages.  The Khan and Lick vanes also use jammed nouns for {\sc ipc}, and Eyre, the {\sc http} vane, now supports {\sc http} subscriptions using jammed nouns in addition to {\sc json}.

@@

\selectlanguage{USenglish}
\printbibliography
\end{document}
